Individual Variation in Animal Movement and Effects on Seed Dispersal Distance

### 2021-02-09 paper/paper.Rmd
discussion

---
title: "Individual Variation in Animal Movement and Effects on Seed Dispersal Distance"
abstract: |
  I don't like this abstract. Revisit at the end...Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an araçari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual araçari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output:
  word_document:
    reference_docx: paper_ref.docx
  html_document:
    df_print: paged
bibliography: mybibfile.bib
---

@@

Even though LDD events are rare or infrequent, they have a disproportionately large effect on gene flow and the genetic pool of populations (Jordano 2017). In the case of well mixed populations, long-distance dispersal can cause random genetic loss through drift or have the opposite effect and maintain high genetic variance in populations that initially drifted (Bohrer, Nathan, and Volis 2005). This is important as in the long-term, these genetic consequences of dispersal can have significant effects for the survival of populations. From a metapopulation perspective, LDD events can have significant consequences for species persistence by enhancing genetic variability and as mechanism for survival in spatially and temporally heterogenous environments. In particular, long-distance dispersal events have a higher probability of reaching isolated populations and therefore establishing a connection and maintaining unrelated populations. It is of special interest to focus in highly heterogenous habitats with rapid change in spatial structure where local extinctions are high, since long-distance dispersal can allow persistence of a metapopulation with immigration and emigration based on long distance dispersal. From a more evolutionary perspective, long-distance dispersal can allow for a species to colonize distant habitats and expand its range, which can also lead to differentiation and speciation (Ronce 2007). We have shown that individual differences in frugivore movement can have significant consequences on the number of long-distance seed dispersal events and it remains to be explored how these individual differences can influence population dynamics in the long term and at larger spatial scales.

# Conclusion

>>>>

---
title: "Individual Variation in Animal Movement and Effects on Seed Dispersal Distance"
abstract: |
  I don't like this abstract. Revisit at the end...Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In the case of frugivore-generated seed dispersal patterns, a large influence is observed by animal movement patterns, and intraspecific variation in animal movement and behaviors can result in interesting seed dispersal patterns. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion and animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an araçari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal with an individual-based model on a homogeneous landscape. This approach showed that there is significant variation in individual araçari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds and estimates for seed dispersal kernel functions. Additionally, we include an approach using statistics of extremes to characterize the tail behavior of generated seed dispersal kernels under varying levels of intraspecific variation of animal movement. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output:
  html_document:
    df_print: paged
  word_document:
    reference_docx: paper_ref.docx
bibliography: mybibfile.bib
---

@@

Frugivore-generated seed dispersal kernels have been used to estimate seed dispersal distances in multiple studies [@jones_closing_2017; @levey_effects_2005; @levey_modelling_2008; @russo_incorporating_2006; @will_mechanistic_2008], often with the goal of understanding how incorporating animal movement and behavior will impact seed dispersal distances. Usually, the focus tends to be on seed dispersal and how spatial patterns of seed aggregation can change in response to animal behavior or landscape heterogeneity. However, fewer studies have emphasized the role of intraspecific variation in animal movement over generated seed dispersal patterns, or even the relative importance that different disperser species may have in carrying seeds and contributing to seed dispersal [@rehm_differences_2018; @zwolak_how_2018]. Extending on the concept of the total dispersal kernel, where we seek to understand and characterize all the different components contributing to a specific parent tree or plant species dispersal kernel, greater emphasis has been given to the role of disperser species and abiotic processes that may influence a total dispersal kernel [@rogers_total_2019]. However, we show that individual variation in animal movement within a single species can also influence a plant's seed dispersal kernel, specifically in a system of toucan-generated seed dispersal kernels. In our particular case, the variation introduced in animal movement patterns included variation at the level of individual animals and variation between family groups. Given that *Pteroglossus pluricinctus* maintain cooperative social groups and previous studies showed overlapping home ranges between individuals belonging to the same social group [@holbrook_home_2011], we wanted to explore how variation in movement between social groups could alter the simulated dispersal kernels. We found that incorporating heterogeneity in animal movement produced seed dispersal kernels with fatter tails, longer dispersal distances, and a higher percentage of long-distance dispersal events compared to null models where no animal movement heterogeneity is included.

Even though long-distance dispersal events are rare or infrequent, they have a disproportionately large effect on gene flow and the genetic pool of populations [@jordano_what_2017]. In the case of well mixed populations, long-distance dispersal can cause random genetic loss through drift or have the opposite effect and maintain high genetic variance in populations that initially drifted [@bohrer2005effects]. This is important as in the long-term, these genetic consequences of dispersal can have significant effects for the survival of populations. From a metapopulation perspective, LDD events can have significant consequences for species persistence by enhancing genetic variability and as mechanism for survival in spatially and temporally heterogenous environments. In particular, long-distance dispersal events have a higher probability of reaching isolated populations and therefore establishing a connection and maintaining unrelated populations. It is of special interest to focus in highly heterogenous habitats with rapid change in spatial structure where local extinctions are high, since long-distance dispersal can allow persistence of a metapopulation with immigration and emigration based on long distance dispersal. From an evolutionary perspective, long-distance dispersal can allow for a species to colonize distant habitats and expand its range [@jordano_what_2017]. We have shown that individual differences in frugivore movement can have significant consequences on the number of long-distance seed dispersal events and it remains to be explored how these individual differences can influence population dynamics in the long term and at larger spatial scales. We also show that the approach using statistics of extremes allows us to estimate the probability of dispersal for distances beyond those directly observed in our simulated dispersal kernels, and that simulation models including individual variation in animal movement have higher probabilities for these events. Our simulated dispersal distances had a maximum of over 1500 meters under the model with individual variation, an estimate that is not unreasonable for this system, and that actually underestimates previous estimates [@holbrook_using_2007]. Parameters estimated for generalized Pareto functions fitted to our frugivore-generated seed dispersal distances showed shape parameters $<0$ pointing towards thin tails constrained by upper bounds. Although we observed differences between the kernels fitted to the three simulation models, all of them still presented parameter values smaller than zero, and thus showing thin tails. Nonetheless, using an approach of statistics of extremes proved to be valuable in selecting seed dispersal thresholds for considering long-distance dispersal events and carrying out analyses focused on the tail of dispersal kernels, whether that is for seeds or pollen [@garcia_extended_2017].

The thing tails estimated by seed dispersal kernel functions can be in part due to the fact that we used a very simple animal movement model incorporating only step lengths and turning angles. Step lengths for all simulation models we randomly drawn from exponential distributions, with varying rate parameters depending on the model. Using an exponential distribution and a simple movement model such as this one can certainly limit the extent to which an animal can move and thus have an upper bound on seed dispersal distances, something we observe not only on fits with the generalized Pareto, but also with the Weibull dispersal kernels. Previous studies looking at a mechanistic understanding of seed dispersal by frugivores have modeled animal movement with approaches similar to this one, varying step lengths to constants or switching to a lognormal distribution [@jones_closing_2017; @levey_effects_2005; @will_mechanistic_2008], or other incorporating more complex animal movement models with diffusion processes[@morales_effects_2006; @pegman_exploring_2017]. There has been emphasis on adding different animal behaviors into mechanistic models of seed dispersal, such as perching or sleeping[@jones_closing_2017; @russo_incorporating_2006], however a focus on individual animal movement is still not coomon. Incorporating individual variation in animal movement models is an active area in the movement ecology field, with various approaches being developed to incorporate this heterogeneity in hierarchical frameworks [@bastille-rousseau_flexible_2016; @borger2012quantifying]. Future directions for this work should focus on taking a step back from seed dispersal, and first characterizing the differences in individual animal movement patterns, since as we observed from our results, individuals generate a wide range of seed dispersal kernels based on their individual movements. In the present study we only focused on using movement rates for each indvidual bird, an average measure of their movement, and thus it is likely that we are also underestimating the frequency of long-distance movements in frugivores. This presents a novel opportunity to also incorporate statistics of extremes for rare long-distance animal movements. Previous work has shown that maximum travel distances for *Pteroglossus pluricinctus* can reach up to 3665 meters in a 30 minute tracking interval [@holbrook_home_2011], estimates that our movement simulations were not able to capture, and thus underestimate how far animals and seeds can travel. In addition to this, our simulations also assume equal contributions of seed dispersal by each individual, when in reality we understand that not all dispersers are the same, with some distributing larger numbers of seeds depending on their breeding status, age or size [don't have a citation for this]. Future research focusing on animal movement models and seed dispersal would also greatly benefit from understanding the relative contributions of individual dispersers to overall seed dispersal kernels.
